---
title: "contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
devtools::load_all(here::here("../vertical-lidar-paper"))
library(GET)
library(patchwork)
library(tibble)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)

ggplot2::theme_set(theme_pub())
alpha <- 0.05 / 3  # (Bonferroni correction)
nperm <- 100  # number of permutations
```

```{r}
# wave can be create once with `source("data-raw/lidar_wave.R")`

# Transform to the GET format
mobs <-                         
  wave %>% 
  dplyr::select(starts_with("ss_")) %>% 
  as.matrix()

cset <- create_curve_set(list(r = seq_len(ncol(mobs)), obs = t(mobs)))

independent_factors <- wave %>% 
  select(
    dominant_species, age_class, crown_closure
  ) %>% 
  mutate(
    dominant_species = factor(dominant_species),
    age_class = factor(age_class)
  ) %>% 
  as.data.frame()
```

## GLM

```{r}
wave_long <- wave %>% 
  pivot_longer(cols = matches("ss_"), names_prefix = "ss_", names_to = "height", names_ptypes = list("height" = integer(0)), values_to = "ss") %>% 
  mutate(
    ss_trans = box_cox(ss + 1e-6)
  )
```

### Diagnose

It seems that issues with residuals could be due to the abundance of near-zero
predictors in some areas of the curve. 
```{r}
wave_long %>% 
  group_by(height, dominant_species) %>% 
  filter(ss < 1e-3) %>% 
  count() %>% 
  ggplot(aes(height / 40 * 100, n / nrow(wave) * 100, fill = dominant_species)) + 
  geom_col() + 
  ylab("Proportion of Y near zero (%)") +
  xlab("Relative height (%)") +
  coord_flip() 
```

# check residuals
```{r}
by_height <- wave_long %>% 
  group_by(height) %>% 
  nest()

box_cox <- function(x) {
  predict(caret::BoxCoxTrans(x + 1e-6), x)
}

sp_mod <- function(x) {
  lm(ss ~ dominant_species + age_class + crown_closure, data = x)
}

sp_mod_sqrt <- function(x) {
  lm(ss_sqrt ~ dominant_species + age_class + crown_closure, data = x)
}

minimum_variance_ratio <- function(x, group, at) {
  # "A rule of thumb is that linear models are fairly robust to heterogeneity
  # of variance so long as the maximum variance is no more than 4× greater
  # than the minimum variance"
  # ---https://stats.stackexchange.com/questions/91872/alternatives-to-one-way-anova-for-heteroskedastic-data/91881#91881
  
  vars <- x %>% 
    group_by({{group}}) %>% 
    summarise(var = var({{at}}, na.rm = TRUE))
  
  max(vars$var) / min(vars$var)
}

by_height <- by_height %>% 
  mutate(
    # boxcox = map(data, ~caret::BoxCoxTrans(.x$ss)),
    # data = map2(data, boxcox, ~mutate(.x, ss_bc = predict(.y, ss))),
    data = map(data, ~mutate(.x, ss_sqrt = sqrt(ss))),
    sp_model = map(data, sp_mod),
    sp_model_sqrt = map(data, sp_mod),
    sp_bptest = map(sp_model, lmtest::bptest),
    sp_bptest_statistic = map_dbl(sp_bptest, "statistic"),
    sp_bptest_p_value   = map_dbl(sp_bptest, "p.value"),
    #sp_ols_bptest = map(sp_model, olsrr::ols_test_breusch_pagan),
    # #sp_levene_test = map(sp_model, ~car::leveneTest(residuals(.x), groups = )),
    # sp_shapiro_test = map(sp_model, ~shapiro.test(sample(residuals(.x), size = 5e3))),
    # sp_shapiro_test_statistic = map_dbl(sp_shapiro_test, "statistic"),
    # sp_shapiro_test_p_value   = map_dbl(sp_shapiro_test, "p.value"),
    # sp_ols_test_f = map(sp_model, ols_test_f),
    
    sp_mvr = map_dbl(data, ~minimum_variance_ratio(.x, dominant_species, ss_sqrt)),
    
    sp_kurtosis = map_dbl(sp_model, ~e1071::kurtosis(residuals(.x))),
    data = map2(data, sp_model, modelr::add_residuals),
    data = map2(data, sp_model, modelr::add_predictions),
    data = map2(data, sp_model, modelr::add_residuals, var = "resid_sqrt"),
    data = map2(data, sp_model, modelr::add_predictions, var = "pred_sqrt")
  )
  


by_height$sp_bptest[[1]]
by_height$sp_ols_bptest[[1]]

by_height %>% 
  ggplot(aes(sp_bptest_statistic, sp_bptest_p_value)) +
  geom_point()

by_height %>% 
  ggplot(aes(sp_kurtosis, sp_shapiro_test_statistic)) +
  geom_point()
```
```{r}
by_height %>%
  filter(height %in% c(1, 24)) %>% 
  unnest(data) %>% 
  #ggplot(aes(pred, resid, color = dominant_species)) +
  ggplot(aes(ss, box_cox(ss), color = dominant_species)) +
  geom_point(size = 0.1, show.legend = FALSE) +
  facet_wrap(~100*height/40, scales = "free")
```

```{r}
# View residuals by group
p <- by_height %>% 
  select(height, data) %>% 
  unnest(data) %>% 
  ggplot(aes(x = height / 40, y = resid, group = stand_id, color = dominant_species)) +
  coord_flip() +
  geom_line(show.legend = FALSE, alpha = 0.5) +
  scale_y_continuous("Scaled lidar return") +
  scale_x_continuous("Scaled height") +
  facet_grid(dominant_species + age_class ~crown_closure)

ggsave("res_fn.pdf", plot = p, width = 12, height = 44, units = "in")

by_height %>% 
  select(height, data) %>% 
  unnest(data) %>% 
  ggplot(aes(x = height / 40, y = resid, group = stand_id, color = dominant_species)) +
  coord_flip() +
  geom_line(show.legend = FALSE, alpha = 0.2)

# View residuals by model (height)
q <- by_height %>% 
  select(height, data) %>% 
  unnest(data) %>% 
  ggplot(aes(x = pred, y = resid, group = NA, color = dominant_species)) +
  geom_point(show.legend = FALSE, size = 0.5, alpha = 0.5) +
  facet_wrap(~I(round(height/40*100)), scales = "free") +
  geom_smooth(se = FALSE)

ggsave("res.pdf", plot = q, width = 12*2, height = 8*2, units = "in")

```

Since there are a lot of zeros in the upper portions, the log transform doesn't
seem to work, and the boxcox transformation either. It might be that the
variance is unequal between groups, which is something we observed. For that
reason, we could try to weight each model based on the inverse of group
variance. Combined to the sqare root transform, it seems to fix a few additional 
heights. 
```{r}
sp_mod_wls <- function(x) {
  lm(ss_sqrt ~ dominant_species + crown_closure + age_class, data = x, weights = weight)
}

by_height <- by_height %>% 
  mutate(
    data = map(data, ~group_by(.x, dominant_species, crown_closure) %>% mutate(weight = 1/var(ss))),
    sp_model_wls = map(data, sp_mod_wls),
    data = map2(data, sp_model_wls, modelr::add_residuals, var = "resid_wls"),
    data = map2(data, sp_model_wls, modelr::add_predictions, var = "pred_wls")
  )

# View residuals by model (height)
q <- by_height %>% 
  select(height, data) %>% 
  unnest(data) %>% 
  ggplot(aes(x = pred, y = resid_wls, group = NA, color = dominant_species)) +
  geom_point(show.legend = FALSE, size = 0.5, alpha = 0.5) +
  facet_wrap(~I(round(height/40*100)), scales = "free") +
  geom_smooth(se = FALSE)

ggsave("res_wls.pdf", plot = q, width = 12*2, height = 8*2, units = "in")
```


The envelopes are created from `r nperm` permutations.

```{r}
functional_anova <- function(g, nsim = nperm, contrasts = TRUE) {
  graph.fanova(
    nsim = nsim,
    curve_set = cset,
    groups = g,
    variances = "unequal",
    test.equality = "var",
    contrasts = contrasts,
    alpha = alpha
  )
}

functional_lm <- function(null_model, nsim = nperm, contrasts = TRUE) {
  graph.flm(
    nsim = nsim,
    formula.full = Y ~ dominant_species + age_class + crown_closure,
    formula.reduced = null_model,
    curve_sets = cset,
    factors = independent_factors,
    contrasts = contrasts,
    GET.args = list(alpha = alpha)
  )
}
res_flm_sp <- functional_lm(Y ~ age_class + crown_closure)
res_flm_cc <- functional_lm(Y ~ dominant_species + age_class, contrasts = FALSE)
res_flm_age <- functional_lm(Y ~ dominant_species + crown_closure, contrasts = FALSE)

res_fanova_sp <- functional_anova(independent_factors$dominant_species)
res_fanova_cc <- functional_anova(independent_factors$crown_closure, contrasts = FALSE)
res_fanova_age <- functional_anova(independent_factors$age_class, contrasts = FALSE)
```

### Contrasts
```{r}
p <- my_plot_contrasts(res_flm_sp, 39, full = TRUE)
pa <- my_plot_contrasts(res_fanova_sp, 39, full = TRUE)
# compare both results (fanova and flm)
p + pa

u <- my_plot_continuous(res_flm_cc, 39, main = "my main")  + scale_y_continuous("Scaled density difference (×10⁻⁴)", labels = function(x) paste0(x * 1000))
v <- my_plot_continuous(res_flm_age, 39) + scale_y_continuous("Scaled density difference (×10⁻³)", labels = function(x) paste0(x * 100))

(p + ggtitle("Species")) +
  ((u + ggtitle("Crown closure")) /
  (v + ggtitle("Age")))

# Save figures independently
# device <- "png" # pdf
# ggsave("figure/fig3-sp-contrasts.png", p, device = device, dpi = "retina", width = 234, height = 234, units = "mm")
# ggsave("figure/fig5-cc-contrasts.png", u, device = device, dpi = "retina", width = 234, height = 234, units = "mm")
# ggsave("figure/fig7-age-contrasts.png", v + scale_y_continuous("breaks = c(-0.2, 0, 0.2)"), device = device, dpi = "retina", width = 234, height = 234, units = "mm")
```

### Summary figure
```{r}
# Species variations ==========================================================
direct_labels <- tribble(
  ~label, ~r, ~curves,
  "Aspen", 0.75, 0.12,
  "Balsam fir", 0.56, 0.09,
  "Paper birch", 0.4, 0.11,
  "White spruce", 0.25, 0.14,
  "Black spruce", 0.05, 0.19
) %>% 
  mutate(a = tolower(label))

sp_v <- 
  res_flm_sp %>% 
  my_plot_data_combined_global_envelope(scale_r = 39) %>% 
  .$ribbon %>% 
  filter(
    type == "Data function",
  ) %>% 
  fill_contrast_matrix() %>% 
  group_by(a, r) %>% 
  summarise(
    curves = sum(abs(curves))
  ) %>% 
  ggplot(aes(x = r, y = curves)) + 
  geom_line(aes(group = a, color = a), show.legend = FALSE) +
  geom_line(data = . %>% group_by(r) %>% summarise(curves = sum(abs(curves)))) +
  geom_text(data = direct_labels, aes(label = label, color = a), show.legend = FALSE, hjust = "left") +
  xlab("Scaled Stand Height") +
  ylab("Sum of absolute scaled density difference") +
  scale_color_brewer("Species", type = "qual", palette = "Dark2") + 
  coord_flip()

# Crown closure variations ====================================================
cc_v <- res_flm_cc %>% 
  my_plot_data_combined_global_envelope(scale_r = 39, main = "main") %>% 
  .$ribbon %>% 
  filter(
    type == "Data function",
  ) %>% 
  fill_contrast_matrix() %>% 
  group_by(a, r) %>% 
  summarise(
    curves = sum(curves)
  ) %>% 
  ggplot(aes(x = r, y = curves)) + 
  geom_line(aes(group = interaction(a), color = a), show.legend = FALSE) +
  geom_line(data = . %>% group_by(r) %>% summarise(curves = sum(abs(curves)))) +
  xlab("Scaled Stand Height") +
  ylab("Sum of scaled density difference") +
  directlabels::geom_dl(aes(label = a, color = a), method = "last.bumpup") +
  scale_color_manual(values = pal_cc) +
  coord_flip()

# Age variations ===============================================================
age_v <- res_flm_age %>% 
  my_plot_data_combined_global_envelope(scale_r = 39) %>% 
  .$ribbon %>% 
  filter(
    type == "Data function",
  ) %>% 
  fill_contrast_matrix() %>% 
  group_by(a, r) %>% 
  summarise(
    curves = sum(curves)
  ) %>% 
  ggplot(aes(x = r, y = curves)) + 
  geom_line(aes(group = interaction(a), color = a), show.legend = FALSE) +
  geom_line(data = . %>% group_by(r) %>% summarise(curves = sum(abs(curves)))) +
  xlab("Scaled Stand Height") +
  ylab("Sum of scaled density difference") +
  scale_color_manual(values = pal_age) +
  directlabels::geom_dl(aes(label = a, color = a), method = "last.bumpup") +
  coord_flip()

sp_v + cc_v + age_v

res_flm_age %>% 
  my_plot_data_combined_global_envelope(scale_r = 39) %>% 
  .$ribbon %>% 
  filter(
    type == "Data function",
  ) %>% 
  fill_contrast_matrix() %>% 
  mutate(
    a = sort_num_factor(a),
    b = sort_num_factor(b)
  ) %>% 
  ggplot(aes(x = r, y = curves, group = interaction(type, a, b))) + 
  geom_ribbon(aes(x = r, ymin = lower, ymax = upper), fill = "black", alpha = 0.01) +
  geom_segment(aes(xend = x_end, yend = y_end, color = ifelse(!inside & !inside_lead, "outside", NA)), show.legend = FALSE) +
  xlab("Scaled Stand Height") +
  ylab("Scaled density difference") +
  facet_grid(~a)+
  coord_flip()
```

