---
title: "contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contrasts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 8
)
```

```{r setup}
devtools::load_all(here::here())
dir.create("figure")
library(GET)
library(patchwork)
library(tibble)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)

op <- options(contrasts = c("contr.sum", "contr.poly"))

ggplot2::theme_set(theme_pub())
save_figures <- function(path, fig, width, height, units = "mm", dpi = "retina", devices = c("pdf", "svg", "png")) {
  lapply(devices, function(dev) ggsave(glue::glue("{path}.{dev}"), fig, device = dev, dpi = dpi, width = width, height = height, units = units))
  invisible(fig)
}
alpha <- 0.05 / 3  # (Bonferroni correction)
nperm <- 3000-1  # number of permutations
m <- 40-1  # number of discrete points per functions
plot_use_trans <- FALSE  # use transformed data for plots

# test variance equality
functional_anova <- function(g, nsim = nperm, contrasts = TRUE, curve_set = cset_trans, ...) {
  GET::graph.fanova(
    nsim = nsim,
    curve_set = curve_set,
    groups = g,
    variances = "unequal",
    test.equality = "var",
    contrasts = contrasts,
    alpha = alpha
    #lm.args = list(weights = weights$weight)
  )
}

# Transformation to equalize variances in groups
corrUnequalVar <- function(curve_set, groups, n.aver = 1L, mirror = FALSE) {
  if (!is.factor(groups)) {
    groups <- factor(groups)
  }
  x <- GET:::data_and_sim_curves(curve_set)
  # Group means, (bar{T}_j(r))
  m <- GET:::groupmeans(x, groups)
  # Sample variance over all functions, Var(T(r))
  varT <- GET:::vvar(x)
  # Variances in the groups, Var(T_j(r))
  v <- GET:::groupvar(x, groups)
  # Moving average
  if(n.aver>1){
    v <- t(apply(v, 1, GET:::maverage, n.aver=n.aver, mirror=mirror))
    varT <- GET:::maverage(varT, n.aver=n.aver, mirror=mirror)
  }
  # Take S_ij(r) = (T_ij(r) - \bar{T}_j(r)) / \sqrt( Var(T_j(r)) ) * \sqrt{Var(T(r))} + \bar{T}_j(r)
  for(i in 1:nrow(x)) {
    x[i,] <- (x[i,] - m[which(rownames(m) == groups[i]),]) / sqrt(v[which(rownames(v) == groups[i]),]) * sqrt(varT) + m[which(rownames(m) == groups[i]),]
  }
  curve_set$funcs <- t(x)
  curve_set
}

wave <- wave %>% 
  mutate(age_class = sort_num_factor(age_class))
```

## Homogeneity

Check homogeneity of variance (Winkler et al. 2014)
```{r}
# wave can be create once with `source("data-raw/lidar_wave.R")`

# Transform to the GET format
mobs <- wave %>% 
  dplyr::select(starts_with("ss_")) %>% 
  as.matrix()

independent_factors <- wave %>% 
  select(
    dominant_species, age_class, crown_closure
  ) %>% 
  mutate(
    dominant_species = factor(dominant_species),
    age_class = factor(age_class)
  ) %>% 
  as.data.frame()

cset <- create_curve_set(list(r = seq_len(ncol(mobs)), obs = t(mobs)))

res_fanova_sp <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset)
my_plot_contrasts(res_fanova_sp, m)

# Successively equalize variance for each group
cset_trans <- corrUnequalVar(cset, groups=with(independent_factors, crown_closure))
res_fanova_sp <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset_trans)
pa <- plot(res_fanova_sp) + coord_flip()

cset_trans <- corrUnequalVar(cset_trans, groups=with(independent_factors, age_class))
res_fanova_sp <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset_trans)
pb <- plot(res_fanova_sp) + coord_flip()

cset_trans <- corrUnequalVar(cset_trans, groups=with(independent_factors, dominant_species))
res_fanova_sp <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset_trans)
pc <- plot(res_fanova_sp) + coord_flip()

pa + pb + pc
```
The order of transformation is important here. From experimentation, we found
that cc>age>sp gives the best results. By using simulated data, we observed that
applying the transformations at once (with e.g. `interaction()`) gives the best
correction because it generally completely removes the heteroscedasticity, while
successive transformations can reintroduce heteroscedasticity in other areas of
the curve. Given our unbalanced data, with missing combinations of predictors,
we could not apply the transformation at once. However, the remaining areas of
heteroscedasticity did not seem to affect the final results of simulated data,
especially with datasets with the same number of observations. The test of
heteroscedasticity ($Z_{ij}(r) = T_{ij}(r) - \bar{T}_j(r)$) appeared to be overly 
powerful for larger datasets (i.e. 5000 observations), than with smaller
datasets (e.g. 200 observations). For example, the test would not indicate
heterogenuous variance on a dataset downsampled from a large dataset in which 
the test found heteroscedasticity.

```{r}
res_fanova_sp <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset) %>% 
  plot() + coord_flip()
res_fanova_sp_trans <- functional_anova(independent_factors$dominant_species, contrasts = FALSE, curve_set = cset_trans) %>% 
  plot() + coord_flip()
res_fanova_cc <- functional_anova(factor(independent_factors$crown_closure), contrasts = FALSE, curve_set = cset) %>% 
  plot() + coord_flip()
res_fanova_cc_trans <- functional_anova(factor(independent_factors$crown_closure), contrasts = FALSE, curve_set = cset_trans) %>% 
  plot() + coord_flip()
res_fanova_age <- functional_anova(independent_factors$age_class, contrasts = FALSE, curve_set = cset) %>% 
  plot() + coord_flip()
res_fanova_age_trans <- functional_anova(independent_factors$age_class, contrasts = FALSE, curve_set = cset_trans) %>% 
  plot() + coord_flip()

# Compare before and after transformation
res_fanova_sp + res_fanova_cc + res_fanova_age
res_fanova_sp_trans + ggtitle("Species") + scale_x_continuous(sec.axis = sec_axis(trans = ~.x/m)) +
  res_fanova_cc_trans + ggtitle("Crown closure") + scale_x_continuous(sec.axis = sec_axis(trans = ~.x/m)) +
  res_fanova_age_trans + ggtitle("Age") + scale_x_continuous(sec.axis = sec_axis(trans = ~.x/m))
```


## GLM

The envelopes are created from `r nperm` permutations.
```{r}
functional_lm <- function(null_model, nsim = nperm, contrasts = TRUE, ...) {
  graph.flm(
    nsim = nsim,
    formula.full = Y ~ dominant_species + age_class + crown_closure,
    formula.reduced = null_model,
    curve_sets = cset_trans,
    factors = independent_factors,
    contrasts = contrasts,
    GET.args = list(alpha = alpha),
    ...
  )
}
res_flm_sp <- functional_lm(Y ~ age_class + crown_closure)
res_flm_cc <- functional_lm(Y ~ dominant_species + age_class, contrasts = FALSE)
res_flm_age <- functional_lm(Y ~ dominant_species + crown_closure, contrasts = FALSE)
```


### p-values
```{r}
p_value(res_flm_sp)
p_value(res_flm_cc)
p_value(res_flm_age)
```

# Figures

Create a transformed set of data with both transformed and untransformed data.
```{r}
trans_ss <- cset_trans[["funcs"]] %>% 
  t() %>%
  as_tibble() %>%
  purrr::set_names(~paste0("trans_", .x))

wave_long <- wave %>%
  bind_cols(trans_ss) %>%
  pivot_longer(
    cols = matches("ss_"),
    names_pattern = "(trans_)?ss_(\\d+)",
    names_to = c("trans", "height"),
    values_to = "ss"
  ) %>% 
  mutate(
        trans = trans != "", 
        height = as.integer(height)
  )
```

## Species
Figure 2. Contrasts from the nonparametric graphical tests of significance for
species. The species in columns are subtracted from the species in rows (e.g.
first column, second row is the aspen - balsam fir contrast). The grey envelope
displays the area of acceptance of the null hypothesis (no differences between
species, p<0.001) from the permutation of the residuals of the null model (eq.
2), the black curve is the average parameter difference (bolder when outside the
envelope). The diagonal displays the `r ifelse(plot_use_trans, "transformed ", "")`
vertical LiDAR return distribution of each species (on a different scale than
the contrasts): black lines represent the median species distribution, shaded
areas represent 95% and 50% variation envelopes.

```{r species_contrasts}
# adapted from: https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-summarise/
quibble_wide <- function(x, q = c(0.025, 0.25, 0.5, 0.75, 0.975)) {
  tibble(x = quantile(x, q), q = q) %>% 
    pivot_wider(1, names_from = q, values_from = x)
}

# Create species figure
psp <- wave_long %>% 
  filter(trans == plot_use_trans) %>% 
  group_by(dominant_species, height) %>% 
  summarise(quibble_wide(ss)) %>% 
  mutate(height = height / m) %>% 
  ggplot(aes(x = height, group = dominant_species)) + 
  geom_ribbon(aes(ymin = `0.025`, ymax = `0.975`), fill = light_grey) + 
  geom_ribbon(aes(ymin = `0.25`, ymax = `0.75`), fill = dark_grey) +
  geom_line(aes(y = `0.5`), colour = "grey20") +
  scale_y_continuous("Density of LiDAR returns", breaks = c(0:2) / 10, expand = c(0, 0)) +
  scale_x_continuous("Relative stand height", limits = c(0, 1), breaks = 1:2/2, expand = c(0, 0)) +
  facet_grid(~dominant_species) + 
  coord_flip(ylim = c(0, 0.30), xlim = c(0, 1))

# Create species contrasts and add species on the diagonal
pcsp <- my_plot_contrasts(res_flm_sp, m, full = TRUE, diag = FALSE) +
  scale_y_continuous(latex2exp::TeX("$\\beta_{m}-\\beta_{n}$")) 
  
save_figures("figure/fig-sp-combined", pcsp, width = 234, height = 234)
```

## Crown closure

Figure 3. Coefficients of the transformed vertical distribution of LiDAR returns
for crown closure. Black lines represent the median distribution, shaded areas
represent 95% and 50% variation envelope. Top right numbers in the panels
indicate the number of observations (stands).
```{r crown_closure}
label_cc <- function(d) paste0(d, "%")
pc_cc <- my_plot_continuous(res_flm_cc, m, main = "my main") + scale_y_continuous(latex2exp::TeX("$\\beta_{cc}$"))

save_figures("figure/fig-glm-cc", pc_cc, width = 234 / 2, height = 234 / 2)
```

Figure 4. `r ifelse(plot_use_trans, "Transformed v", "V")`ertical distribution 
of LiDAR returns as a function of crown closure and species. Black lines
represent the median distribution, shaded areas represent 95% and 50% variation
envelope. Top right numbers in the panels indicate the number of observations
(stands).
```{r crown_closure_contrasts}
pcc <- wave_long %>% 
  filter(trans == plot_use_trans) %>% 
  group_by(dominant_species, height, crown_closure) %>% 
  summarise(quibble_wide(ss), .groups = "drop_last") %>% 
  mutate(height = height / m) %>% 
  ggplot(aes(x = height, group = interaction(dominant_species, crown_closure))) + 
  geom_ribbon(aes(ymin = `0.025`, ymax = `0.975`), fill = light_grey) + 
  geom_ribbon(aes(ymin = `0.25`, ymax = `0.75`), fill = dark_grey) +
  geom_line(aes(y = `0.5`), colour = "grey20") +
  geom_vline(aes(xintercept = 0)) +
  scale_x_continuous("Relative stand height", limits = c(0, 1), breaks = 1:2/2, expand = c(0, 0)) +
  scale_y_continuous("Density of LiDAR returns", breaks = c(0:2) / 10, expand = c(0, 0)) +
  facet_grid(dominant_species~label_cc(crown_closure)) +
  coord_flip(ylim = c(0, 0.25), xlim = c(0, 1)) +
  ( if(plot_use_trans) 
    # transformed data
    list(
      coord_flip(ylim = c(-0.1, 0.2), xlim = c(0, 1)),
      scale_y_continuous(breaks = c(-2, 0, 2, 4, 6) / 10, expand = c(0, 0))
      )
  )
pcc

save_figures("figure/fig-cc", pcc, width = 234, height = 234)
```

## Age
Figure 5. `r ifelse(plot_use_trans, "Transformed v", "V")`ertical distribution 
of LiDAR returns as a function of stand density and species. Black lines
represent the median distribution, shaded areas represent 95% and 50% variation
envelope. Top right numbers in the panels indicate the number of observations
(stands).
```{r age}
label_age <- function(x) paste(x, "yr")
# upgrade labels
attr(res_flm_age, "labels") <- label_age(attr(res_flm_age, "labels"))
page <- my_plot_continuous(res_flm_age, m) +
  coord_flip(ylim = c(-0.05, 0.15), expand = c(0, 0)) +
  scale_y_continuous(latex2exp::TeX("$\\beta_{age}$"), labels = c(-5, 0, 5, 10) / 100, breaks = c(-5, 0, 5, 10) / 100)
page

save_figures("figure/fig-age", page, width = 234, height = 234 / 2)
```

Figure 6. `r ifelse(plot_use_trans, "Transformed v", "V")`ertical distribution 
of LiDAR returns as a function of age groups and species. Black lines
represent the median distribution, shaded areas represent 95% and 50% variation
envelope. Top right numbers in the panels indicate the number of observations
(stands).
```{r age_sp}
page2 <- wave_long %>% 
  filter(trans == plot_use_trans) %>% 
  group_by(dominant_species, height, age_class) %>% 
  summarise(quibble_wide(ss), .groups = "drop_last") %>% 
  mutate(height = height / m) %>% 
  ggplot(aes(x = height, group = interaction(dominant_species, age_class))) + 
  geom_ribbon(aes(ymin = `0.025`, ymax = `0.975`), fill = light_grey) + 
  geom_ribbon(aes(ymin = `0.25`, ymax = `0.75`), fill = dark_grey) +
  geom_line(aes(y = `0.5`), colour = "grey20") +
  geom_hline(aes(yintercept = 0), color = "grey50", size = 0.1) +
  facet_grid(dominant_species~sort_num_factor(label_age(age_class))) +
  scale_x_continuous("Relative stand height", limits = c(0, 1), breaks = 1:2/2, expand = c(0, 0)) +
  scale_y_continuous("Density of LiDAR returns", breaks = c(0:2) / 10, expand = c(0, 0)) +
  coord_flip(ylim = c(-0, 0.25), xlim = c(0, 1)) +
  ( if(plot_use_trans)
    # transformed data
    list(
      coord_flip(ylim = c(-0.1, 0.25), xlim = c(0, 1)),
      scale_y_continuous("Density of LiDAR returns (transformed)", breaks = c(0, 10, 20) / 100, expand = c(0, 0))
      )
  )
page2

save_figures("figure/fig-age2", page2, width = 234, height = 234)
```


# Appendix

```{r}
cap <- paste(
  "Each line is an observation, the bold line is the median function",
  "of the group, the grey ribbon displays the 95% and 50% intervals.",
  "The number in the top right corner is the number of observations."
)

wl <- wave_long %>% filter(!trans) # use untransformed data
papu <- wl %>% 
  group_by(dominant_species, crown_closure, age_class, height) %>% 
  summarise(
    n = n(),
    median = median(ss),
    q25 = quantile(ss, probs = 25/100),
    q75 = quantile(ss, probs = 75/100),
    q025 = quantile(ss, probs = 2.5/100),
    q0975 = quantile(ss, probs = 97.5/100)
    ) %>% 
  ggplot(aes(x = height / m, y = median, group = interaction(dominant_species, crown_closure, age_class))) +
  coord_flip(ylim = c(0, 0.25)) +
  geom_line(data = wl, aes(y = ss, color = dominant_species, group = stand_id), show.legend = FALSE, size = 0.1, alpha = 0.4) +
  geom_line(aes(color = dominant_species), show.legend = FALSE) +
  geom_ribbon(aes(ymin = q025, ymax = q0975), fill = "#000000", alpha = 0.2) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#000000", alpha = 0.2) +
  scale_y_continuous("Density of LiDAR returns", breaks = c(0:2) / 10, expand = c(0, 0)) +
  scale_x_continuous("Relative stand height", limits = c(0, 1), breaks = 1:2/2) +
  geom_text(data = wl %>% 
              distinct(dominant_species, crown_closure, age_class, stand_id) %>% 
              count(dominant_species, crown_closure, age_class), 
            aes(label = n), x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 2.5) +
  facet_grid(dominant_species + age_class ~ crown_closure) +
    labs(
    title = "Vertical distribution of LiDAR returns",
    subtitle = "Figure A-1", 
    caption = wrapper(cap, 80)
    )

wlt <- wave_long %>% filter(trans) # use untransformed data
papt <- wlt %>% 
  group_by(dominant_species, crown_closure, age_class, height) %>% 
  summarise(
    n = n(),
    median = median(ss),
    q25 = quantile(ss, probs = 25/100),
    q75 = quantile(ss, probs = 75/100),
    q025 = quantile(ss, probs = 2.5/100),
    q0975 = quantile(ss, probs = 97.5/100)
    ) %>% 
  ggplot(aes(x = height / m, y = median, group = interaction(dominant_species, crown_closure, age_class))) +
  coord_flip(ylim = c(0, 0.25)) +
  geom_line(data = wlt, aes(y = ss, color = dominant_species, group = stand_id), show.legend = FALSE, size = 0.1, alpha = 0.4) +
  geom_line(aes(color = dominant_species), show.legend = FALSE) +
  geom_ribbon(aes(ymin = q025, ymax = q0975), fill = "#000000", alpha = 0.2) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#000000", alpha = 0.2) +
  scale_y_continuous("Density of LiDAR returns (transformed)", breaks = c(0:2) / 10, expand = c(0, 0)) +
  scale_x_continuous("Relative stand height", limits = c(0, 1), breaks = 1:2/2) +
  geom_text(data = wlt %>% 
              distinct(dominant_species, crown_closure, age_class, stand_id) %>% 
              count(dominant_species, crown_closure, age_class), 
            aes(label = n), x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 2.5) +
  facet_grid(dominant_species + age_class ~crown_closure) +
  labs(
    title = "Vertical distribution of LiDAR returns (transformed)",
    subtitle = "Figure A-2", 
    caption = wrapper(cap, 80)
    )

ggsave("figure/appendix-1.pdf", papu + theme_pub(10), width = 6, height = 22, units = "in")
ggsave("figure/appendix-2.pdf", papt + theme_pub(10), width = 6, height = 22, units = "in")
```

## Formal test of homoscedasticity of resdiuals using BP
```{r breusch-pagan}
full_mod <- function(x) {
  lm(ss ~ dominant_species + age_class + crown_closure, data = x)
}

sp_mod <- function(x) {
  lm(ss ~ age_class + crown_closure, data = x)
}
cc_mod <- function(x) {
  lm(ss ~ dominant_species + age_class, data = x)
}
age_mod <- function(x) {
  lm(ss ~ dominant_species + crown_closure, data = x)
}

my_bp <- function(data, model) {
  # Use the model formula to create the Breusch-Pagan Test using the squared
  # residuals. Little hack: Substitue the left-hand side of the formula with the
  # squared residuals unevaluated (the data must contain a .resid column).
  stopifnot(".resid" %in% names(data))
  f <- model$call[[2]]
  f[[2]] <- quote(I(.resid^2))
  lm(f, data = data)
}

add_mod <- function(x, mod) {
  mutate(x,
         model = map(data, mod),
         aug = map(model, broom::augment),
         data = map2(data, aug, ~bind_cols(.x, .y %>% select(.resid))),
         resid_mod = map2(data, model, my_bp),  # full
         summary = map(resid_mod, broom::glance),
         bptest = map(model, lmtest::bptest),
         bptest_statistic = map_dbl(bptest, "statistic"),
         bptest_p_value   = map_dbl(bptest, "p.value"),
  ) %>% 
    unnest(summary)
}

plot_mod <- function(x) {
  qplot(bptest_statistic, height / m, data = x) + geom_path() +  # this plot just confirms the bp values with the R² (×N)
  qplot(r.squared, height / m, data = x) + geom_path() +
  qplot(bptest_p_value, height / m, data = x) + geom_path() + geom_vline(xintercept = 5/100, alpha = 0.5, size = 2, color = "blue") + geom_text(aes(label = round(height * 100 / m)))
}
  

wlt_by_height <- nest(wlt, data = -height)


ds <- function(..., data, n = 100) {
  #set.seed(seed)
  #slice_sample(data, n = n) %>% 
    lm(..., data = data)
}
seed <- sample.int(9999, 1)

# Full model
wlt_by_height %>% 
  add_mod(full_mod) %>% 
  plot_mod()

# Reduced models:
wlt_by_height %>% 
  add_mod(sp_mod) %>% 
  plot_mod()

wlt_by_height %>% 
  add_mod(cc_mod) %>% 
  plot_mod()

wlt_by_height %>% 
  add_mod(age_mod) %>% 
  plot_mod()

# Untransformed data (highly significant, and max R²~0.2
wl %>% 
  nest(data = -height) %>% 
  add_mod(full_mod) %>% 
  plot_mod()
```

## R-squared

We can compute the R-squared for each $h$ of every model. 
```{r rsq}
add_mod_rsq <- function(x, mod, col) {
  mutate(x,
         model = map(data, mod),
         glance = map(model, broom::glance),
         rsq = map_dbl(glance, "adj.r.squared")
  ) 
}

wlt_by_height <- nest(wlt, data = -height)

# Full model R²
r_squared <- wlt_by_height %>% 
  add_mod_rsq(full_mod) %>% 
  mutate(
    r_squared_full = rsq
  ) %>% 
  # sp
  add_mod_rsq(sp_mod) %>% 
  mutate(
    r_squared_dominant_species = rsq
  )  %>% 
  # cc
  add_mod_rsq(cc_mod) %>% 
  mutate(
    r_squared_crown_closure = rsq
  ) %>% 
  # age
  add_mod_rsq(age_mod) %>% 
  mutate(
    r_squared_age_class = rsq
  )

prsq <- r_squared %>% 
  select(-(data:rsq)) %>% 
  pivot_longer(-height, names_to = "model", names_prefix = "r_squared_", values_to = "r_squared") %>% 
  mutate(
    model_ordered = factor(model, levels = c("dominant_species", "crown_closure", "age_class", "full")),
    model_label = forcats::fct_recode(model_ordered, "Species" = "dominant_species", "Crown closure" = "crown_closure", "Age" = "age_class", "Full" = "full")
  ) %>% 
  ggplot(aes(height / m, r_squared, group = model)) +
  geom_path(aes(color = model_ordered), show.legend = FALSE) +
  geom_text(data = . %>% 
              filter(
                  model == "age_class" & height == 14 |
                  model == "crown_closure" & height == 23 |
                  model == "full" & height == 28 |
                  model == "dominant_species" & height == 33 
                  ) %>% 
              mutate(
                r_squared = r_squared + c(7, 0, 5, 10) / 100
              ), aes(label = model_label, color = model), hjust = 1.2, vjust = 1, show.legend = FALSE) +
  coord_flip() +
  scale_x_continuous("Relative stand height", expand = c(0, 1/100)) +
  scale_y_continuous(latex2exp::TeX("$R^2$"), limits = c(0, 0.48), expand = c(0, 0)) +
  scale_color_manual("Model", values = pal_4)
prsq

save_figures("figure/fig-r-squared", prsq, width = 234 / 2, height = 234)
```

