---
title: "contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contrasts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

```{r setup}
devtools::load_all(here::here("../vertical-lidar-paper"))
library(GET)
library(patchwork)
library(tibble)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)

ggplot2::theme_set(theme_pub())
alpha <- 0.05 / 3  # (Bonferroni correction)
nperm <- 100  # number of permutations
```

The observations are imbalanced. The most important imbalance is between 
species, then age. The crown closures are more balanced, but still not 
perfectly balanced. 

## Diagnostic

What's the magnitude of the problem and how are the differences between the age 
and crown closure classes combined? Here I'll use the filtered data.
```{r}
# read data
# wave can be create once with `source("data-raw/lidar_wave.R")`
wave <- wave %>%
    mutate(
    age_class = sort_num_factor(age_class),
    crown_closure = sort_num_factor(crown_closure)
)
```


```{r}
group_size <- wave  %>% 
  group_by(dominant_species, age_class, crown_closure) %>% 
  tally(name = "count") %>% 
  group_by(dominant_species) %>% 
  add_count(name = "total") 

group_size %>% 
  ggplot(aes(crown_closure, age_class, fill = count / total, label = count)) +
  geom_tile(show.legend = FALSE) +
  scale_fill_continuous(trans = "log") +
  facet_wrap(~dominant_species) +
  geom_text(color = "grey80", size = 3) +
  coord_equal()
```

Aspen is the smallest category, with very little observations in each category. 
Also balsam fir has substantially more observations than any other species. 
Another imbalance is that some species are more represented in some classes. 
Black and white spruces have lot of young stands, while paper birch stands are 
mostly mature. I would expect the null model to account for these differences. 

## Downsampling

Downsample observations to mimic Aspen distribution for species comparison.
```{r}
set.seed(2020)
ec_bf_down_aspen <- wave %>% 
  group_by(dominant_species, age_class, crown_closure) %>% 
  nest() %>% 
  left_join(
    group_size %>% 
      ungroup() %>% 
      filter(dominant_species == "aspen") %>% 
      select(-dominant_species),
    by = c("crown_closure", "age_class")
  ) %>% 
  mutate(
    count = replace_na(count, 0),
    data = map2(data, count, ~sample_n(.x, round(.y), replace = TRUE))
  ) %>% 
  unnest(data) %>% 
  ungroup()

# visualize results
ec_bf_down_aspen  %>% 
  group_by(dominant_species, age_class, crown_closure) %>% 
  tally(name = "count") %>% 
  group_by(dominant_species) %>% 
  add_count(name = "total") %>% 
  mutate(
    count = ifelse(count < 1, NA, count)
  ) %>% 
  ggplot(aes(crown_closure, age_class, fill = count, label = count)) +
  geom_tile(show.legend = FALSE) +
  scale_fill_continuous(trans = "log") +
  facet_wrap(~dominant_species) +
  geom_text(color = "grey80", size = 3) +
  coord_equal()
```



### Contrasts

We setup the analysis
```{r}
mobs <- ec_bf_down_aspen %>% 
  dplyr::select(starts_with("ss_")) %>% 
  as.matrix()

cset <- create_curve_set(list(r = 1:39, obs = t(mobs)))

independent_factors <- ec_bf_down_aspen %>% 
  select(
    dominant_species, age_class, crown_closure
  ) %>% 
  as.data.frame()
```

```{r fig.width = 8, fig.height = 8}
functional_lm <- function(null_model, nsim = nperm, contrasts = TRUE) {
  graph.flm(
    nsim = nsim,
    formula.full = Y ~ dominant_species + age_class + crown_closure,
    formula.reduced = null_model,
    curve_sets = cset,
    factors = independent_factors,
    contrasts = contrasts,
    GET.args = list(alpha = alpha)
  )
}

res_flm_sp <- functional_lm(Y ~ age_class + crown_closure)
res_flm_cc <- functional_lm(Y ~ dominant_species + age_class, contrasts = FALSE)
res_flm_age <- functional_lm(Y ~ dominant_species + crown_closure, contrasts = FALSE)
```


```{r fig.width = 8, fig.height = 8}
my_plot_contrasts(res_flm_sp, 39, full = TRUE)
```

My understanding is that I should balance the other factors for each analysis 
(i.e. balance crown closure per cc class, balance age per age class), but let's 
see what it does to other comparisons.

```{r fig.width = 8, fig.height = 4}
my_plot_continuous(res_flm_cc, 39) + scale_y_continuous("Scaled density difference (×10⁻³)", labels = function(x) paste0(x * 100))
my_plot_continuous(res_flm_age, 39)  + scale_y_continuous("Scaled density difference (×10⁻³)", labels = function(x) paste0(x * 100))
```

It seems that age is very much affected because aspen only has three age classes 
sufficently represented (30, 50, 70), crown closure is less affected  because it
is better represented, but the differences are still smaller than with the 
complete data.

## Upsampling

Another strategy is to upsample our observations by padding the data with
observations drawn at random. Each missing observation becomes a penalty, since 
it is replaced by a random observation from the global data.

```{r}
set.seed(2020)
target_size <- 30

wave_upsample <- wave %>% 
  # pad empty classes
  expand(dominant_species, age_class, crown_closure) %>%  
  left_join(wave, by = c("dominant_species", "age_class", "crown_closure")) %>% 
  # sample
  group_by(dominant_species, age_class, crown_closure) %>% 
  nest() %>% 
  mutate(
    count = target_size,
    # # Permutation only (nothing exits the envelope)
    # data = map2(data, count, ~sample_n(wave %>% select(matches("ss_")), round(.y), replace = TRUE))
    
    # # Pick within the available observations for the specific combination
    # data = map2(data, count, ~sample_n(.x, round(.y), replace = TRUE))
    
    # If sufficient observations, sample a fraction of the data
    # If missing observations, randomly pick observations from the pool
    data = map2(data, count, ~sample_n_if_needed(.x %>% drop_na(), round(.y), fill = wave))
  ) %>% 
  unnest(data) %>% 
  ungroup()

# visualize results
wave_upsample  %>% 
  group_by(dominant_species, age_class, crown_closure) %>% 
  tally(name = "count") %>% 
  group_by(dominant_species) %>% 
  add_count(name = "total") %>% 
  mutate(
    count = ifelse(count < 1, NA, count)
  ) %>% 
  ggplot(aes(crown_closure, age_class, fill = count, label = count)) +
  geom_tile(show.legend = FALSE) +
  scale_fill_continuous(trans = "log") +
  facet_wrap(~dominant_species) +
  geom_text(color = "grey80", size = 3)
```

Setup the functional format
```{r}
mobs <- wave_upsample %>% 
  dplyr::select(starts_with("ss_")) %>% 
  as.matrix()

cset <- create_curve_set(list(r = 1:39, obs = t(mobs)))

independent_factors <- wave_upsample %>% 
  select(
    dominant_species, age_class, crown_closure
  ) %>% 
  as.data.frame()

res_flm_sp <- functional_lm(Y ~ age_class + crown_closure)
res_flm_cc <- functional_lm(Y ~ dominant_species + age_class, contrasts = FALSE)
res_flm_age <- functional_lm(Y ~ dominant_species + crown_closure, contrasts = FALSE)
```

Species
```{r fig.width = 8, fig.height = 8}
my_plot_contrasts(res_flm_sp, 39, full = TRUE)
```

```{r fig.width = 8, fig.height = 4}
(
  my_plot_continuous(res_flm_cc, 39)  + 
   scale_y_continuous("Scaled density difference (×10⁻³)", labels = function(x) paste0(x * 100)) + 
    ggtitle("Crown closure")
   ) / 
(
  my_plot_continuous(res_flm_age, 39) + 
    scale_y_continuous("Scaled density difference (×10⁻³)", labels = function(x) paste0(x * 100)) +
    ggtitle("Age"))
```
